Preventing a Nuclear ISIL: \\Recommendations for United States Policy

### 2015-05-11 draft.tex
Updated Policy Chart coloration, removed an unnecessary map, added in Prussin & Nacht joke references.

@@

@@

@@

@@

@@

@@

\includepdf[angle=270, pagecommand={\thispagestyle{fancy2}\null\vfill\captionof{figure}{Non-linear Logic Tree}\label{fig:logic_tree}}, scale=0.85]{./figures/NucSec_Logic_Model2.pdf}

@@

\begin{figure}[H]
 \centering
 \includegraphics[trim = 0cm 0cm 0cm 0cm, clip,scale=0.55]{./figures/Policy_Chart3.png}
   \caption{Summary of existing and proposed policies to combat ISIL acquisition of nuclear weapons.}
     \label{fig:policy_chart}
\end{figure}

@@

\begin{figure}[H]
 \centering
 \includegraphics[trim = 0cm 0cm 0cm 0cm, clip,scale=0.5]{./figures/current_territory.png}
   \caption{Current ISIL Territory \cite{BBC2015a}}
     \label{fig:current_territory}
\end{figure}

\begin{figure}[H]
 \centering
 \includegraphics[trim = 0cm 0cm 0cm 0cm, clip,scale=0.3]{./figures/current2.png}
   \caption{What ISIL Wants \cite{Lewis2014}}
     \label{fig:current2}
\end{figure}

@@

While avoiding unintended consequences or wasted resources are compelling reasons to reconsider the proposed policies, perhaps the most compelling reason is to place the threat and proposed policies into context.  It is perhaps a bit generous to even call ISIL obtaining a nuclear weapon a low probability event as there is zero basis in history for it to be able to do so \cite{pollack2014unthinkable,Bracken2013,Reed2010,Agency2004}. Developing nuclear weapons amidst crippling sanctions, airstrikes, and a dedicated opposition has never been done, despite several nation states attempting to do so \cite{Reed2010,Agency2004}.  In fact, the only nation to develop a nuclear weapon in the time of war was the U.S., and the U.S. mainland was never seriously threatened. Supposing that ISIL was the proverbial \enquote{Black Swan} case, it is worth noting that, with the exception of the U.S., no state has gone from decision to having a deliverable nuclear weapon in under four years \cite{Reed2010}. Even modern conservative estimates place the timeline at 5-6 years for a competent nation state pursuing an optimal path to a nuclear weapon \cite{Harney2006}. Proponents of the threat of nuclear terrorism will likely point to the possibility that ISIL will circumvent the indigenous development route completely and pursue theft or purchase of key components or a complete device.  Again, there is no history to suggest that a nation state would take that risk given the enormous repercussions it would face \cite{pollack2014unthinkable}.  Additionally, anything short of a complete device would still require significant development to weaponized, perhaps three to four years depending on the amount of assistance ISIL received \cite{Harney2006}.

@@

Typical nuclear reactions release energies on the order of 1-10 MeV \cite{Prussin2014}. In comparison, typical chemical reactions release energy on the order of a single eV per molecule, and perfectly combusted TNT, a common high-explosive, releases approximately 35 eV per molecule \cite{Prussin2014}. Clearly the amount of energy released in nuclear reactions is many orders of magnitude greater than typical chemical reactions more familiar to society. However, the energy released in fission, the primary nuclear reaction in nuclear weapons, is 100 times that of typical nuclear reactions (approximately 180 MeV per reaction) \cite{krane1987introductory,Loveland2005}. Combining the incredible energy released per fission reaction with the density of the fuel in nuclear weapons, it becomes readily apparent that nuclear weapons contain an energy source greater than any other on earth. Combined  with the incredibly short timescales on which nuclear reactions occur (on the order of 10\(^{-7}\) s - 10\(^{-6}\) s),  nuclear weapons become tools of unimaginable power and destruction \cite{Loveland2005,Cochran1994}.

>>>>

FPCON - Force Protection Condition

@@

LEU - Low Enriched Uranium

@@

NFAA - Nuclear Forensics and Attribution Act

@@

NPR - Nuclear Posture Review

@@

OIR - Operation Inherent Resolve

@@

SOCOM - U.S. Special Operations Command

@@

\includepdf[angle=270, pagecommand={\thispagestyle{fancy2}\null\vfill\captionof{figure}{ISIL Logic Tree}\label{fig:logic_tree}}, scale=0.85]{./figures/NucSec_Logic_Model2.pdf}

@@

\begin{figure}[H]
 \centering
 \includegraphics[trim = 0cm 0cm 0cm 0cm, clip,scale=0.52]{./figures/PolicyChart4.png}
   \caption{Summary of existing and proposed policies to combat ISIL acquisition of nuclear weapons.}
     \label{fig:policy_chart}
\end{figure}

@@

\begin{figure}[H]
 \centering
 \includegraphics[trim = 0cm 0cm 0cm 0cm, clip,scale=0.5]{./figures/current_territory.png}
   \caption{Current ISIL Territory \cite{BBC2015a,Lewis2014}}
     \label{fig:current_territory}
\end{figure}

% \begin{figure}[H]
%  \centering
%  \includegraphics[trim = 0cm 0cm 0cm 0cm, clip,scale=0.3]{./figures/current2.png}
%    \caption{What ISIL Wants \cite{Lewis2014}}
%      \label{fig:current2}
% \end{figure}

@@

While avoiding unintended consequences or wasted resources are compelling reasons to reconsider the proposed policies, perhaps the most compelling reason is to place the threat and proposed policies into context.  It is perhaps a bit generous to even call ISIL obtaining a nuclear weapon a low probability event as there is zero basis in history for it to be able to do so \cite{pollack2014unthinkable,Bracken2013,Nacht1867,Reed2010,Agency2004}. Developing nuclear weapons amidst crippling sanctions, airstrikes, and a dedicated opposition has never been done, despite several nation states attempting to do so \cite{Reed2010,Agency2004}.  In fact, the only nation to develop a nuclear weapon in the time of war was the U.S., and the U.S. mainland was never seriously threatened. Supposing that ISIL was the proverbial \enquote{Black Swan} case, it is worth noting that, with the exception of the U.S., no state has gone from decision to having a deliverable nuclear weapon in under four years \cite{Reed2010}. Even modern conservative estimates place the timeline at 5-6 years for a competent nation state pursuing an optimal path to a nuclear weapon \cite{Harney2006}. Proponents of the threat of nuclear terrorism will likely point to the possibility that ISIL will circumvent the indigenous development route completely and pursue theft or purchase of key components or a complete device.  Again, there is no history to suggest that a nation state would take that risk given the enormous repercussions it would face \cite{pollack2014unthinkable}.  Additionally, anything short of a complete device would still require significant development to weaponized, perhaps three to four years depending on the amount of assistance ISIL received \cite{Harney2006}.

@@

Typical nuclear reactions release energies on the order of 1-10 MeV \cite{Prussin2014}. In comparison, typical chemical reactions release energy on the order of a single eV per molecule, and perfectly combusted TNT, a common high-explosive, releases approximately 35 eV per molecule \cite{Prussin2014}. Clearly the amount of energy released in nuclear reactions is many orders of magnitude greater than typical chemical reactions more familiar to society. However, the energy released in fission, the primary nuclear reaction in nuclear weapons, is 100 times that of typical nuclear reactions (approximately 180 MeV per reaction) \cite{krane1987introductory,Loveland2005}. Combining the incredible energy released per fission reaction with the density of the fuel in nuclear weapons, it becomes readily apparent that nuclear weapons contain an energy source greater than any other on earth. Combined  with the incredibly short timescales on which nuclear reactions occur (on the order of 10\(^{-7}\) s - 10\(^{-6}\) s),  nuclear weapons become tools of unimaginable power and destruction \cite{Loveland2005,Cochran1994,Prussin2015}.
